---
title: "Perceptions of drinking water management and safety: A public survey of U.S. residents"

author: 
  - name: T. Allen Berthold
    email: allen.berthold@ag.tamu.edu
  - name: Audrey McCrary
    email: audrey.mccrary@ag.tamu.edu
    orcid: 0000-0002-7061-6113
  - name: Stephanie deVilleneuve
    orcid: 0009-0002-3984-035X
    email: s.devilleneuve@ag.tamu.edu
  - name: Michael Schramm
    orcid: 0000-0003-1876-6592
    email: michael.schramm@ag.tamu.edu

bibliography: refs.bib
csl: https://www.zotero.org/styles/council-of-science-editors-author-date?source=1




---


```{r}
#| label: setup
#| echo: false
#| message: false
#| warning: false

library(distributional)
library(ggdist)
library(modelsummary)
library(ragg)
library(survey)
library(targets)
library(tidyverse)
library(tinytable)

knitr::opts_chunk$set(dev = "ragg_png",
                      echo = FALSE,
                      fig.width = 6,
                      fig.asp = 0.618,
                      dpi = 300)

font_hoist <- function(family, silent = FALSE) {
  font_specs <- systemfonts::system_fonts() |> 
    dplyr::filter(family == {{family}}) |> 
    dplyr::mutate(family = paste(family, style)) |> 
    dplyr::select(plain = path, name = family)
  
  purrr::pwalk(as.list(font_specs), systemfonts::register_font)
  
  if (!silent)  message(paste0("Hoisted ", nrow(font_specs), " variants:\n",
                               paste(font_specs$name, collapse = "\n")))
}

font_hoist("Manrope")

```




Texas Water Resources Institute, Texas A&M AgriLife, 
College Station Texas, USA (Berthold, McCrary, deVilleneuve, Schramm). 
Correspondence to Schramm: michael.schramm@ag.tamu.edu)

**Research Impact Statement:** TBD

**ABSTRACT:** TBD

(**Keywords:**  drinking water; public perception; water quality; survey.)

\newpage

# Introduction


# Methodology

## Survey Instrument

The online survey was administered and distributed through the Qualtrics online
panel in April 2023. Panels were recruited using gender, age, and 
race/ethnicity quotas to obtain approximately representative samples of the U.S.
population. Panels were continuously sampled until a representative sample was
obtained, with a total sample size of 1,100 respondent and estimated ±3% margin
of error at 95% confidence [@dillmanMailInternetSurveys2011]. Quality checks,
performed by Qualtrics, excluded responses that failed attention or speed checks, 
IP address checks, or failed to meet representative demographic criteria. 
All participants were provided and agreed to a written informed consent agreement.
The Texas A&M University Institutional Review Board reviewed the study
protocol prior to distribution and deemed the study exempt from formal review.

We sought to characterize (1) choice in primary drinking water source; (2)
problems experienced with drinking water; (3) concern about drinking water
contamination; (4) perceptions of trust and safety in drinking water quality
and associated management entities; and (5) how people receive information
about drinking water quality. Participants were asked for their main source
of household tap water (municipal public supply, public rural water district,
private ground or surface water, private rainwater collection) and their main
source of drinking water (unfiltered tap, filtered tap, bottled/prepackaged, 
or other). We also asked if they had experienced any of
12 different potential drinking water issues. To characterize concerns about
contamination sources, we also asked how concerned they were about 10 different 
potential sources of contamination on a five-point scale from 
*"not at all concerned"* to *"extremely concerned."* 

Levels of responsibility and trust in different entities involved in drinking
water management were assessed with 2 questions. First, participants were asked
what level of responsibility (on a five-point scale from *"no responsibility"* 
to *"full responsibility"*) they believed each of six different entities had for
ensuring drinking water is safe. We also asked what the participants level of
trust was in each of those entities (on a five-point scale from *"do not trust at all"* 
to *"fully trust"*). To assess the public's perceived safety of drinking 
water, we asked participants to rate their drinking water safety on an ordinal 
scale from 0 through 10. We assessed sources of drinking water quality 
information by asking how often (five-point scale from *"never"* to *"always"*)
each of 11 different sources were used to get information about drinking water
quality. Questions are documented in the supplementary information (Table S1).

## Analysis

Survey responses were weighted to match marginal sex/gender, age group, 
race/ethnicity, and education proportions to national level benchmarks from the 
5-year 2021 American Community Survey (ACS; Table S2).
We chose to weight gender using "male" and "non-male" binary responses because
the ACS only provides binary responses for sex. 
This approach groups "female" and "other" respondents under the same marginal 
weight. 
Race and ethnicity were recoded at white or Caucasian and non-white categories 
due to small subpopulation sample sizes. 
Weights were developed by poststratification raking using the American National Election 
Study (ANES) weighting algorithm implemented in the *anesrake* R package 
[@debellComputingWeightsAmerican2009; @pasekAnesrakeANESRaking2018].
Weighted summary statistics were calculated using the *survey* R package
[@lumleyAnalysisComplexSurvey2004]. 
Confidence intervals for proportions
were computed using the "logit" method with the "svyciprop" function.



## Models

We explored factors associated with changes in (1) primary drinking water source,
(2) perceived safety of drinking water, and (3) perception of drinking water 
safety using regression-based approaches.
Each of our models included demographic variables (Table S2), source
of household tap water, and experience with any of (1) odd taste, (2) 
odd smell, (3) discolored water, or (4) cloudy water as explanatory
variables. 
Primary drinking water source (unfiltered tap water, 
filtered tap water, or bottled water) was modeled using a multinomial 
logistic regression model [@agrestiCategoricalDataAnalysis2002].
We used proportional odds models 
[@mccullaghRegressionModelsOrdinal1980; @agrestiCategoricalDataAnalysis2002] to 
explore factors associated (1) trust in entities responsible for drinking 
water safety (five-point scale), and (2) self-reported rating (on scale 1-10) 
of drinking water safety. Multinomial logistic regressions were fit using 
the *svyVGAM* R package [@lumleySvyVGAM2023] and proportional odds models were 
fit using the *survey* R package [@lumleyAnalysisComplexSurvey2004].
We calculated the generalized variance inflation factor (GVIF) to evaluate 
potential impacts of multi-collinearity among model predictors on our 
interpretation of the models [@foxGeneralizedCollinearityDiagnostics1992].
We report our model results as odds-ratios.





# Results

## Sources of drinking water and drinking water problems

The majority of respondents lived in homes with a municipal public water supply 
(56%, 1.6% SE) with 13.3% using some type of private water supply (Table \@ref(tab:q5q6)).
A relatively high percentage of respondents stats they did not know what the 
source of their tap water was (13.3%, 1.2% SE). 
The majority of respondents (65.5%) used unfiltered or filtered tap water as 
their main source of drinking water (Table \@ref(tab:q5q6)). Bottled or pre-packaged 
water was the primary drinking water sources for 34.2% (1.6% SE) of respondents.
Only 3.1% (0.6% SE) of respondents stated that they had not experienced any
issues with their tap water (Table \@ref(tab:q7)). Odd taste (27.5%, 1.4% SE), hardness 
or scale buildup (20.3%, 1.3% SE), and odd smell (18.8%, 1.3% SE) were the most 
common issues. Notably, 14.9% (1.1% SE) of respondents stated they have 
experienced a boil water notice.

```{r}
#| label: q5q6
#| tab.cap: The proportion of respondents with the specified household water source and drinking water source.
#| results: asis

label <- knitr::opts_current$get('label')
caption <- knitr::opts_current$get('tab.cap')

tar_read(q5_q6_summary) |> 
  print("markdown") |> 
  capture.output() |> 
  replace(3, values = "+:==========================================+===============:+=======:+") |> 
  (function(.) c(paste0("Table: (\\#tab:", label, ") ", caption, ")"),
                 (""),
                 .))() |> 
  cat(sep = "\n")

```





```{r}
#| label: q7
#| tab.cap: Proportion of respondents that have experienced specified issues with their drinking water.
#| results: asis

label <- knitr::opts_current$get('label')
caption <- knitr::opts_current$get('tab.cap')


tar_read(q7_summary) |> 
  print("markdown") |> 
  capture.output() |> 
  replace(4, values = "+:=====================================+===============:+=======:+") |> 
  (function(.) c(paste0("Table: (\\#tab:", label, ") ", caption, ")"),
                 .))() |> 
  cat(sep = "\n")

```


```{r}
#| label: q11
#| fig.cap: Contaminants of concern


tar_read(q11_summary)$p1
```

We did not observe strong differences among response levels between different
contaminants of potential concern (Figure \@ref(fig:q11)). Pesticides (*extremely 
concerned* = 21%, 1.3% SE; *very concerned* = 21.1%, 1.3% SE), industrial 
chemicals (*extremely concerned* = 20.3%, 1.3% SE; *very concerned* = 20.1%, 
1.3% SE), pathogens (*extremely concerned* = 19.9%, 1.3% SE; *very concerned* = 
20.8%, 1.3% SE), and heavy metals (*extremely concerned* = 19.8%, 1.3% SE; 
*very concerned* = 22.0%, 1.3% SE) are sources that elicited the highest
levels of concern across all respondents. Disinfection byproducts 
(*Not at all concerned* = 22.6%, 1.4% SE; *slightly concerned* = 20.7%, 1.3% SE) 
and PFAS (*not at all concerned* = 23.0%, 1.4% SE; *slightly concerned* = 17.8%, 
1.2% SE) were sources of least concern.



```{r}
#| label: m2
#| fig-cap: Primary drinking water model coefficients (odds-ratios) and 95% CIs. Odds-ratios are relative to unfiltered tap water.
#| fig-asp: 1.1
tar_read(m2_coefs)$p1
```

Based on the multinomial regression model, we found limited effects of 
demographic variables, drinking water sources, and experience with odd taste, 
color, or smell on the odds of using bottled drinking water compared to
unfiltered drinking water (Figure \@ref(fig:m2), Table S3). Notably, females had 1.49 times the odds 
(95% CI [1.02, 2.17]) of using bottled drinking water instead of unfiltered 
tap water compared to males. Non-whites had 1.59 times the odds 
(95% CI [1.01, 2.52]) of using bottled drinking water instead of unfiltered 
drinking water compared to white respondents. Renters had 1.64 times the odds
(95% CI [1.07, 2.50]) compared to homeowners.

Conversely, older respondents were more likely to use unfiltered drinking water
compared to filtered water (45:54 age group: OR = 0.29, 95% CI [0.132, 9.636];
55:64 age group: OR = 0.29, 95% CI [0.130, 0.647]; 65+ age group: OR = 0.278, 
95% CI [0.002, 1.25]) but there was not strong evidence for differences
in odds between bottled water and unfiltered drinking water among any of the 
age groups. Education also showed strong effects on the use of filtered drinking
water. Having a bachelor's degree resulted in 4.31 times the odds of using 
filtered drinking water compared to those not graduating high school.
A master's degree had a similarly large effect (OR =  3.06, 95% CI [1.09, 8,59]).

## Sources of information


```{r}
#| label: q12
#| fig-cap: Sources of drinking water quality information.


tar_read(q12_summary)$p1
```

Water providers appear to be the most relied upon source for information on
drinking water with 13.9% (1.1% SE) and 16.4% (1.2% SE) of responses indicating
water providers are *always* or *often* used as a source of information 
(Figure \@ref(fig:q12)). Both social media (*always* = 11.5%, ±1% SE; *often* = 15.2%, ±1.1% SE) 
and word-of-mouth (*always* = 9.4%, ±1% SE; *often* = 17.6%, ±1.2% SE) are also frequent
sources of information with similar or higher usage than local news agencies 
(*always* = 8.7%, ±0.9% SE; *often* = 14.3%, ±1.1% SE) or local newspapers
(*always* = 6.6%, ±0.8% SE; *often* = 15.3%, ±1.1% SE). Some of the least
utilized sources included radio or podcasts (*never* = 40.5%, ±1.6% SE; 
*occasionally* = 20.1%, ±1.3% SE), email (*never* = 46.7%, 
±1.6% SE; *occasionally* = 19.6%, ±1.3% SE), national newspapers (*never* = 
51.8%, ±1.6% SE; *occasionally* = 16.7%, ±1.2% SE), and government agencies 
(*never* = 37.5%, ±1.6% SE; *occasionally* = 22.4%, ±1.4% SE).




## Perceived responsibility, trust, and safety

```{r}
#| echo: false

## function to grab and print values from figure 4 and 5 data
paste_trust_vals <- function(df, question, response, paren = FALSE, ...) {
  
  label_fn <- scales::label_percent(...)
  
  proportion <- df |> 
    filter(Question == question,
           Response == response) |> 
    pull(proportion) |> 
    label_fn()
  
  low <- df |> 
    filter(Question == question,
           Response == response) |> 
    pull(proportion_low) |> 
    label_fn()
  
  high <- df |> 
    filter(Question == question,
           Response == response) |> 
    pull(proportion_upp) |> 
    label_fn()
  
  if(!isTRUE(paren)){
    paste0(proportion, ", 95% CI [", low, ", ", high, "]")
  } else {
    paste0(proportion, " (95% CI [", low, ", ", high, "])")
  }
    
}

paste_trust_vals_source <- function(df, question, response, q5, paren = FALSE, ...) {
  
  label_fn <- scales::label_percent(...)
  
  proportion <- df |> 
    filter(Question == question,
           Response == response,
           Q5 == q5) |> 
    pull(proportion) |> 
    label_fn()
  
  low <- df |> 
    filter(Question == question,
           Response == response,
           Q5 == q5) |> 
    pull(proportion_low) |> 
    label_fn()
  
  high <- df |> 
    filter(Question == question,
           Response == response,
           Q5 == q5) |> 
    pull(proportion_upp) |> 
    label_fn()
  
  if(!isTRUE(paren)){
    paste0(proportion, ", 95% CI [", low, ", ", high, "]")
  } else {
    paste0(proportion, " (95% CI [", low, ", ", high, "])")
  }
    
}
```


```{r}
#| label: q14
#| fig-cap: Weighted response percentages for the level of responsibility people believe different entities have for ensuring drinking water safety. Points represent the population-level estimated percent response, error bars are the 95% confidence intervals.
#| fig.asp: 2
#| out.width: "50%"

tar_read(q14_summary)$p1

```

`r paste_trust_vals(tar_read(q14_summary)$df, "Water Utility","Full responsibility",paren=TRUE,accuracy = 0.1)` 
of people said that water utilities have *"full responsibility"* for ensuring 
drinking water safety (Figure \@ref(fig:q14)).
Local and state government also received high percentages of *"full responsibility"*
by the public 
(local government = `r paste_trust_vals(tar_read(q14_summary)$df, "Local Government","Full responsibility",accuracy = 0.1)`;
state government = `r paste_trust_vals(tar_read(q14_summary)$df, "State Government","Full responsibility",accuracy = 0.1)`).
Relative to other governmental institutions, fewer people felt the federal
government had full responsibility 
(`r paste_trust_vals(tar_read(q14_summary)$df, "Federal Government","Full responsibility",accuracy = 0.1)`) 
for drinking water safety but a relatively higher percentage indicated
the federal government has *"moderate responsibility"* 
(`r paste_trust_vals(tar_read(q14_summary)$df, "Federal Government","Moderate responsibility",accuracy = 0.1)`).
Moderate responsibility was the most common response for the level of 
responsibility attributed to household residents 
(`r paste_trust_vals(tar_read(q14_summary)$df, "Residents","Moderate responsibility",accuracy = 0.1)`) 
and landlords/property managers 
(`r paste_trust_vals(tar_read(q14_summary)$df, "Landlord/property manager","Moderate responsibility",accuracy = 0.1)`).
A relatively high proportion of responses indicated residents have no 
responsibility for drinking water quality 
(`r paste_trust_vals(tar_read(q14_summary)$df, "Residents","No responsibility",accuracy = 0.1)`).


```{r}
#| label: q15
#| fig-cap: Weighted response percentages for the level of trust people have in the entities responsible for ensuring drinking water safety. Points represent the population-level estimated percent response, error bars are the 95% confidence intervals.
#| fig.asp: 2
#| out.width: "50%"

tar_read(q15_summary)$p1

```


Water utilities were more trusted by the public than other governmental
institutions (Figure \@ref(fig:q15)). We found a higher percentage of respondents
 *"fully trust"* water utilities 
(`r paste_trust_vals(tar_read(q15_summary)$df, "Water Utility","Fully trust",accuracy = 0.1)`) 
for making sure that drinking water is safe compared to state 
(`r paste_trust_vals(tar_read(q15_summary)$df, "State Government","Fully trust",accuracy = 0.1)`), 
local (`r paste_trust_vals(tar_read(q15_summary)$df, "Local Government","Fully trust",accuracy = 0.1)`), 
and the federal government 
(`r paste_trust_vals(tar_read(q15_summary)$df, "Federal Government","Fully trust",accuracy = 0.1)`).
Water utilities and local government had higher rates of *"mostly trust"* responses
(water utility = `r paste_trust_vals(tar_read(q15_summary)$df, "Water Utility","Mostly trust",accuracy = 0.1)`,
local government = `r paste_trust_vals(tar_read(q15_summary)$df, "Local Government","Mostly trust",accuracy = 0.1)`)
relative to state government (`r paste_trust_vals(tar_read(q15_summary)$df, "State Government","Mostly trust",accuracy = 0.1)`) and federal government (`r paste_trust_vals(tar_read(q15_summary)$df, "Federal Government","Mostly trust",accuracy = 0.1)`).

The federal government received the highest rates of distrust with 
`r paste_trust_vals(tar_read(q15_summary)$df, "Federal Government","Do not trust at all", paren = TRUE, accuracy = 0.1)`
of responses indicating they *"do not trust at all"* the federal government
for ensuring safety of their drinking water (Figure \@ref(fig:q15)). 
State and local governments fared slightly better with distrust, with 
`r paste_trust_vals(tar_read(q15_summary)$df, "State Government","Do not trust at all",paren = TRUE,accuracy = 0.1)`
and `r paste_trust_vals(tar_read(q15_summary)$df, "Local Government","Do not trust at all",paren = TRUE,accuracy = 0.1)`
of responses stating they *"do not trust at all"* state and local government respectively
for ensuring safety of drinking water.
Water utilities fared the best among governmental institutions for distrust with
`r paste_trust_vals(tar_read(q15_summary)$df, "Water Utility","Do not trust at all",paren = TRUE,accuracy = 0.1)` 
indicating they *"do not trust at all"* water utilities for ensuring safety of 
drinking water.

Landlords/property managers received rates of trust and distrust that appear
very similar to response rates for the federal government 
(*"fully trust"* = `r paste_trust_vals(tar_read(q15_summary)$df, "Landlord/property manager","Fully trust",accuracy = 0.1)`;
 *"do not trust at all"* = `r paste_trust_vals(tar_read(q15_summary)$df, "Landlord/property manager","Do not trust at all",accuracy = 0.1)`). 
 Comparatively, household residents received the highest responses rates for 
 *"fully trust"* 
 (`r paste_trust_vals(tar_read(q15_summary)$df, "Residents","Fully trust",accuracy = 0.1)`) 
 and lowest responses rates for  *"do not trust at all"*
 (`r paste_trust_vals(tar_read(q15_summary)$df, "Residents","Do not trust at all",accuracy = 0.1)`)
 among all entities.
The marginal response rates by primary tap water source (Figure S1) do not
suggest that there are strong differences between those on private water sources
or public water sources when it comes to trust in the ability of residents to ensure drinking water is safe. 
<!-- Note, if we look at the mariginal response rate by -->
<!-- primary tap water source, there is not much evidence that -->
<!-- there are differences in trust and distrust in resident ensuring water quality -->
<!-- for those with private supplies or public supplies. -->


<!-- Don't need this, it is in the model dummy  -->
<!-- We also examined rates of trust and distrust by the respondent's primary tap  -->
<!-- water source (Figure S1). A larger percentage of municipal water customers  -->
<!-- indicated they *"fully trust"* (`r paste_trust_vals_source(tar_read(q15_by_source)$df, "Water Utility", "Fully trust", "Public supply - municipal", accuracy = 0.1)`)  -->
<!-- or *"mostly trust"*  (`r paste_trust_vals_source(tar_read(q15_by_source)$df, "Water Utility", "Mostly trust", "Public supply - municipal", accuracy = 0.1)`) water utilities for -->
<!-- providing safe drinking water compared to people with private water supplies  -->
<!-- (*"fully trust"* = `r paste_trust_vals_source(tar_read(q15_by_source)$df, "Water Utility", "Fully trust", "Private supply - well, river, pond, rainwater", accuracy = 0.1)`; -->
<!-- *"fully trust"* = `r paste_trust_vals_source(tar_read(q15_by_source)$df, "Water Utility", "Mostly trust", "Private supply - well, river, pond, rainwater",  accuracy = 0.1)`) -->


```{r}
#| label: m1-m3
#| fig-cap: Estimated effect sizes (odds-ratios) and 95% confidence intervals of selected variables on (1) level of trust in drinking water utilities, and (2) perceived safety of drinking water.
#| fig-asp: 1.1
#| 
tar_read(m1_m3_coefs)
```





The proportional odds model provides evidence that sex/gender, home ownership,
household tap supply, and experience with taste/odor/color issues are associated
with the perception of drinking water safety (Figure \@ref(fig:m1-m3), Table S4). 
Females were `r round(1/0.76, 2)` times more likely to rate their drinking
water less safe than males (OR = 0.76, 95% CI [0.59, 0.97]).
Respondents identifying their gender as other had `r round(1/0.13, 2)` times 
the odds (OR = 0.13, 95% CI [0.05, 0.32]) of rating their drinking water less safe
than male respondents.
Home ownership provided a strong signal with renters `r round(1/0.67,2)` times 
more likely to provide a lower drinking water safety ranking than homeowners 
(OR = 0.67, 95% CI [0.51, 0.89]).
Respondents on private water supplies had 1.54 times that odds of providing
a higher drinking water safety rating than people on municipal supplies 
(OR = 1.54, 95% CI [1.10, 2.16]). 
People that experienced taste/odor/color issues with `r round(1/0.32, 2)` times
more likely to provide a lower safety rating than those that never experienced 
an issue (OR = 0.32, 95% CI [0.24, 0.41]).

We identified difference among some age groups, but a clear pattern was not
evident. Respondents in the 45-54 age bracket were `r round(1/0.45, 2)`
times more likely to provide a lower drinking water safety rating than the
18-24 year old age group.



# Discussion

# Conclusion


# Supporting Information

Additional supporting information may be found online under the Supporting 
Information tab for this article: Tables with survey questions, unweighted and 
weighted survey demographics, summary tables for regression models and 
additional figures of marginal survey responses.

# Data Availability

Data and code used in analysis is publicly available in TBD.

# Acknowledgements

TBD

# Literature Cited