---
title: "Perceptions of Drinking Water and Safety: A Public Survey of U.S. Residents"
output: github_document
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(survey)
library(svyVGAM)
tidy.svyVGLM <- function(
    x, 
    conf.int = FALSE, 
    conf.level = 0.95,
    exponentiate = FALSE, 
    ...
){
  # Replace `summary(x)$coefficients` with `summary(x)$coeftable`
  s <- summary(x)
  s <- s$coeftable
  ret <- tibble::as_tibble(s, rownames = "term")
  
  # All of this stays the same:
  colnames(ret) <- c("term", "estimate", "std.error", "statistic", "p.value")
  coefs <- tibble::enframe(stats::coef(x), name = "term", value = "estimate")
  ret <- dplyr::left_join(coefs, ret, by = c("term", "estimate"))
  if (conf.int){
    ci <- broom:::broom_confint_terms(x, level = conf.level, ...)
    ret <- dplyr::left_join(ret, ci, by = "term")
  }
  if (exponentiate){ret <- broom:::exponentiate(ret)}
  
  ret
  
}
```

This is the data and code repository associated with 
the draft manuscript, "Perceptions of Drinking Water and Safety: A Public Survey of U.S. Residents"
currently in submission. Authors include: Karissa Palmer, T. Allen Berthold, Audrey McCrary, Stephanie deVilleneuve, Michael Schramm, and Holli R. Leggette



This project uses the [renv](https://rstudio.github.io/renv/) and [targets](https://docs.ropensci.org/targets/) R packages to facilitate reproducibility. Once the project is downloaded and
opened in RStudio install the renv package:

```{r install, eval=FALSE}
install.packages("renv")
```

The packages and package versions used in the project can be installed and restored using:
```{r restore, eval=FALSE}
renv::restore()
```

To reproduce the analysis you will need a Census API key which can be obtained from https://api.census.gov/data/key_signup.html. Then use targets to rerun the analysis. This will take a little while to run:
```{r make, eval=FALSE}
tidycensus::census_api_key("API KEY GOES HERE")
targets::tar_make()
```


Survey responses used in our analysis are stored in the `clean_survey` target:

```{r data}
targets::tar_read(clean_survey)
```


## Models


### RQ1

Research Question 1 was inferred using a weighted proportional odds model:
$$
\ln \left( \frac{P(Y\leq j)}{P(Y>j)} \right) = \alpha_j - \beta_1 x_1 - \cdots - \beta_n x_n 
$$

where the left hand side of the equation is the log odds of the cumulative probability
of an outcome being at up to and including level $j$. The outcome levels are
ordinal rating 1,...,10 in response to the question: "How safe do you believe your
drinking water is for consumption?"
Predictor variables included sex/gender, age category, race/ethnicity, education, homeownership, income, primary tap water supply, and experience with any of (1) odd taste, (2) odd smell, (3) discolored water, or (4) cloudy water.


The RQ1 model is stored in target `m1`:

```{r m1}
gtsummary::tbl_regression(targets::tar_read(m1), exponentiate = TRUE) |> 
  gtsummary::as_gt() |> 
  gt::as_raw_html()
```

### RQ2

Research question 2 asks how much trust does the public have in drinking water entities.
Again we fit a proportion-odds model to ordinal rating levels
(No responsibility, Some responsibility, Moderate responsibility, Most responsibility, Full Responsibility)
in response to the question, "What level of responsibility should the following entities
have for making sure drinking water is safe?"
Our predictor variable of interest are drinking water entities: Federal Government,
State government, Local government, Water utility, Landlord/property manager, and household residents.

Results for this model are stored in target `m4`:

```{r m4}
gtsummary::tbl_regression(targets::tar_read(m4), exponentiate = TRUE) |> 
  gtsummary::as_gt() |> 
  gt::as_raw_html()
```

A second model was fit to ordinal rating levels (Do not trust at all, Somewhat trust, Moderately trust, Mostly trust, Fully trust) in response to the questions, "What is your level of trust in the following entities for making sure your drinking water is safe for consumption?" Our predictor variable of interest are drinking water entities. Results are stored in target `m5`.

```{r m5}
gtsummary::tbl_regression(targets::tar_read(m5), exponentiate = TRUE) |>   
  gtsummary::as_gt() |> 
  gt::as_raw_html()
```

Next we explored predictors associated with changes in the probability of
trust ratings for drinking water utilities specifically. Here we fit the 
proportional odds model to responses of trust level for drinking water utilities
with sex/gender, age category, race/ethnicity, education, homeownership, income, primary tap water supply, and experience with any of (1) odd taste, (2) odd smell, (3) discolored water, or (4) cloudy water as predictors. Results are stored in target `m3`

```{r m3}
gtsummary::tbl_regression(targets::tar_read(m3), exponentiate = TRUE) |> 
  gtsummary::as_gt() |> 
  gt::as_raw_html()
```

### RQ3

We used a weighted multinomial regression model to explore the effects of 
predictor variables on choices in primary drinking water source (unfiltered tap water, filtered tap water, and bottled water).
Demographic variables, tap water sources, and experience with taste/odor/color issues were
the independent variables. Results are stored in target `m2`:

```{r m2}
m2 <- targets::tar_read(m2)
gtsummary::tbl_regression(m2, exponentiate = TRUE) |> 
  gtsummary::as_gt() |> 
  gt::as_raw_html()
```