---
title: "Perceptions of drinking water management and safety: A public survey of U.S. residents"
abstract: "tbd"
pub-date:
  - title: Submitted
    date: "`datetime(year: 2024, month: 05, day: 01)`{=typst}"
author:
  - name: T. Allen Berthold
    affiliations: 1
  - name: Audrey McCrary
    orcid: 0000-0002-7061-6113
    affiliations: 1
  - name: Michael Schramm
    orcid: 0000-0003-1876-6592
    email: michael.schramm@ag.tamu.edu
    affiliations: 1
  - name: Stephanie deVilleneuve
    orcid: 0009-0002-3984-035X
    affiliations: 1
institute:
  - id: 1
    name: Texas Water Resources Institute, Texas A&M AgriLife
format:
  quarto-lapreprint-typst:
    papersize: us-letter
    columns: 1
    bibliography: refs.bib
    bibliographystyle: apa
    mainfont: "Minion Pro"
kind: "Draft Preprint"

knitr: 
  opts_chunk: 
    echo: false
    fig-width: 6
    fig-asp: 0.618
    out-width: "100%"
    fig-dpi: 300
    dev: "ragg_png"

---

```{r}
#| label: setup
#| echo: false
#| message: false
#| warning: false

library(distributional)
library(ggdist)
library(modelsummary)
library(ragg)
library(survey)
library(targets)
library(tidyverse)
library(tinytable)

font_hoist <- function(family, silent = FALSE) {
  font_specs <- systemfonts::system_fonts() |> 
    dplyr::filter(family == {{family}}) |> 
    dplyr::mutate(family = paste(family, style)) |> 
    dplyr::select(plain = path, name = family)
  
  purrr::pwalk(as.list(font_specs), systemfonts::register_font)
  
  if (!silent)  message(paste0("Hoisted ", nrow(font_specs), " variants:\n",
                               paste(font_specs$name, collapse = "\n")))
}

font_hoist("Manrope")

```

# Introduction

Dramatic declines in mortality rates since the late 1800s are largely attributed
to advances and adoption of clean water technologies 
[@cutlerRolePublicHealth2005].
However, even in the United States (U.S.), there is clear evidence of unequal access 
to safe drinking water and variable trust towards the entities responsible for providing 
clean drinking water [@meehanExposingMythsHousehold2020].
Recent high profile drinking water crises in U.S. cities (notably Flint, 
Michigan and Jackson, Mississippi) highlight the intersection of water insecurity,
racial injustices, and distrust in authorities responsible for 
delivering safe drinking water [@meehanExposingMythsHousehold2020; @pauliFlintWaterCrisis2020; @mengUrbanWaterCrisis2023].
The combination of regulatory/utility management system failures with historical
power imbalances results in diminished trust by marginalized communities with 
public drinking water systems and the entities responsible for ensuring safety 
[@rosingerHomeTapWater2020; @jaffeeUnequalTrustBottled2024].

Characterizing consumer perceptions of drinking water safety across the U.S. is
a challenge due to the diversity in community composition, experiences, 
drinking water sources, and fragmentation in governance and responsibility of
drinking water safety. Community water systems (CWS) provide water for 
household use to approximately 283 million Americans, while another 42.5 million 
Americans utilize self-supplied sources such as private wells [@dieterEstimatedUseWater2018].
The size, capacity, and population served by CWSs are highly variable. CWSs serving
less than 10,000 people account for 91% of CWSs, but serve only 16% of the U.S.
population; while the largest 9% of CWS provide water to 83% of the U.S. population 
[@levinUSDrinkingWater2024]. Among CWSs, 61% of withdrawals are from surface
water but self-supplied water is overwhelmingly (98%) sourced from groundwater 
[@dieterEstimatedUseWater2018].

On a national scale, drinking water provided through public utilities is 
considered safe based on compliance metrics set out by the U.S. Environmental
Protection Agency, with 7-8% of CWSs reporting at least one violation annually 
[@allaireNationalTrendsDrinking2018].
However, the occurrence of these violations fall hardest on marginalized communities.
The body of evidence shows that small and rural CWSs, especially those serving
disadvantaged communities have higher likelihoods of drinking water violations 
[@levinUSDrinkingWater2024; @mcdonaldDrinkingWaterViolations2018; 
@allaireNationalTrendsDrinking2018; @baeEthnicityPovertyRace2023; @strosniderRuralUrbanDifferences2017; @switzerClassRaceEthnicity2018].
Increased rates of violations among small CWSs include limited technical
capacity, smaller labor markets, reduced economies of scale, and economic 
constraints preventing investments in aging infrastructure 
[@levinUSDrinkingWater2024; @switzerHumanCapitalResource2016; @scottDrinkWhatYou2018]. 
While less frequent, large systems are not immune from failure to provide safe 
drinking water as demonstrated in Flint, Jackson, and other municipalities. 
These high profile failures lead to confirmation of preexisting distrust of 
government institutions by communities that have experienced marginalization 
[@jaffeeUnequalTrustBottled2024; @pierceMistrustTapFactors2017]. 
Tap water avoidance has consistently increased
among children, Hispanics, non-Hispanic black, and lower socioeconomic status 
communities since the mid 2015 Flint Michigan crisis, suggesting decreased
levels of trust in tap water [@rosingerHomeTapWater2020].
Furthermore, emerging contaminants, such as per- and 
polyfluorinated substances, present new challenges to drinking water systems and
the failure to rapidly address these contaminants further erodes trust
in drinking water safety [@levinUSDrinkingWater2024; 
@domingoHumanExposurePolyfluoroalkyl2019; @bertholdLetTalkPFAS2023].
Underscoring perceptions about drinking water safety are people's trust
in the institutions and public authorities responsible for providing safe water
[@teodoroProfitsDistrustCitizenconsumers2022].
While public trust in government institutions is at historically low levels 
[@pewresearchcenterAmericansViewsGovernment2022], there is some limited 
evidence that the public may have higher trust in local 
public water utilities relative to other state, local, and federal 
institutions [@weisnerComplexitiesTrustUrban2020]. 
Confidence and trust in public drinking water and local utilities can
be improved through meaningful engagement, campaigns, and intervention programs
[@lawmanHydratePhillyIntervention2020; @pierceMistrustTapFactors2017; 
@paragBattleBottlesBuilding2009].

The limited research on self-supplied water users indicates that private well
users are more likely to perceive their water as safe relative to people
on public water systems [@gholsonConsumerWaterQuality2018; @huBottledWaterUnited2011].
In the U.S., private wells are not regulated under the Safe Drinking Water Act
or other rules to ensure safety/quality of the water and the 
management of those wells are left to the owner. Therefore, private well programs
offered through local health departments or state extension services play
an important role in educating well owners on the need for maintenance and water
quality testing to ensure drinking water safety [@waitDisparitiesWellWater2020; 
@gholsonConsumerWaterQuality2018]. 
Evidence suggests rates of tap water 
contamination can be relatively high among private wells 
[@pieperIncidenceWaterborneLead2015; @knobelochPrivateDrinkingWater2013; 
@swistockWaterQualityManagement2013; @craunCausesOutbreaksAssociated2010] 
despite the belief in high water quality.
Furthermore, relatively few well owner conduct regular testing or maintenance
activities [@kreutzwiserUnderstandingStewardshipBehaviour2011; @gholsonConsumerWaterQuality2018]
and the educational support for these activities can be highly variable [@waitDisparitiesWellWater2020].

Decreased trust in water providers or water quality results in the increased
reliance and usage of alternative sources of drinking water such as bottled 
water [@jaffeeUnequalTrustBottled2024; @huBottledWaterUnited2011]. 
Evidence also points to factors such as age, gender, race/ethnicity, and 
dissatisfaction with taste/color with increased reliance on bottled water 
sources [@doriaBottledWaterTap2006; @gorelickPerceptionsWaterIncreased2011; 
@huBottledWaterUnited2011; @rosingerHomeTapWater2020].
Increased reliance on more expensive bottled water source further contribute
to issues of water insecurity in already marginalized communities.


*note* - need to link above to research questions:

We used a national opt-in online survey to examine (1) public perceptions of
drinking water safety, (2) public trust in entities involved in drinking water 
management, (3) problems experienced with drinking water, (4) concern about
specific source of contamination, and (5) relied upon source of drinking water 
information.


# Methodology

## Survey Instrument

The online survey was administered and distributed through the Qualtrics online
panel in April 2023. Panels were recruited using gender, age, and 
race/ethnicity quotas to obtain approximately representative samples of the U.S.
population. Panels were continuously sampled until a representative sample was
obtained, with a total sample size of 1,100 respondent and estimated ±3% margin
of error at 95% confidence [@dillmanMailInternetSurveys2011]. Quality checks,
performed by Qualtrics, excluded responses that failed attention or speed checks, 
IP address checks, or failed to meet representative demographic criteria. 
All participants were provided and agreed to a written informed consent agreement.
The Texas A&M University Institutional Review Board reviewed the study
protocol prior to distribution and deemed the study exempt from formal review.

We sought to characterize (1) choice in primary drinking water source; (2)
problems experienced with drinking water; (3) concern about drinking water
contamination; (4) perceptions of trust and safety in drinking water quality
and associated management entities; and (5) how people receive information
about drinking water quality. Participants were asked for their main source
of household tap water (municipal public supply, public rural water district,
private ground or surface water, private rainwater collection) and their main
source of drinking water (unfiltered tap, filtered tap, bottled/prepackaged, 
or other). We also asked if they had experienced any of
12 different potential drinking water issues. To characterize concerns about
contamination sources, we also asked how concerned they were about 10 different 
potential sources of contamination on a five point scale from 
*"not at all concerned"* to *"extremely concerned."* 

Levels of responsibility and trust in different entities involved in drinking
water management were assessed with 2 questions. First, participants were asked
what level of responsibility (on a five point scale from *"no responsibility"* 
to *"full responsibility"*) they believed each of six different entities had for
ensuring drinking water is safe. We also asked what the participants level of
trust was in each of those entities (on a five point scale from *"do not trust at 
all"* to *"fully trust*"). To assess the public's perceived safety of drinking 
water, we asked participants to rate their drinking water safety on an ordinal 
scale from 0 through 10. We assessed sources of drinking water quality 
information by asking how often (five point scale from *"never"* to *"always*")
each of 11 different sources were used to get information about drinking water
quality. Questions are documented in the supplementary information (Table S1).

## Analysis

Survey responses were weighted to match marginal sex/gender, age group, 
race/ethnicity, and education proportions to 
national level benchmarks from the 5-year 2021 American Community Survey (ACS).
We chose to weight gender using "male" and "non-male" binary responses because
the ACS only provides binary responses for sex. This approach groups "female" and
"other" respondents under the same marginal weight. Race and ethnicity were 
recoded at white or Caucasian and non-white categories due to small subpopulaiton
sample sizes. Weights were 
developed by poststratification raking using the American National Election 
Study (ANES) weighting algorithm implemented in the *anesrake* R package 
[@debellComputingWeightsAmerican2009; @pasekAnesrakeANESRaking2018].
Weighted summary statistics were calculated using the *survey* R package
[@lumleyAnalysisComplexSurvey2004].



## Models

We explored factors associated with changes in (1) primary drinking water source,
(2) perceived safety of drinking water, and (3) perception of drinking water 
safety using regression-based approaches.
Each of our models included demographic variables (Table S1), source
of household tap water, and experience with any of (1) odd taste, (2) 
odd smell, (3) discolored water, or (4) cloudy water as explanatory
variables. 
Primary drinking water source (unfiltered tap water, 
filtered tap water, or bottled water) was modeled using a multinomial 
logistic regression model [@agrestiCategoricalDataAnalysis2002].
We used proportional odds models 
[@mccullaghRegressionModelsOrdinal1980; @agrestiCategoricalDataAnalysis2002] to 
explore factors associated (1) trust in entities responsible for drinking 
water safety (five-point scale), and (2) self-reported rating (on scale 1-10) 
of drinking water safety. Multinomial logistic regressions were fit using 
the *svyVGAM* R package [@lumleySvyVGAM2023] and proportional odds models were 
fit using the *survey* R package [@lumleyAnalysisComplexSurvey2004].
We calculated the generalized variance inflation factor (GVIF) to evaluate 
potential impacts of multi-collinearity among model predictors on our 
interpretation of the models [@foxGeneralizedCollinearityDiagnostics1992].
We report our model results as odds-ratios.





# Results

## Sources of drinking water and drinking water problems

The majority of respondents lived in homes with a municipal public water supply 
(56%, 1.6% SE) with 13.3% using some type of private water supply ([@tbl-q5q6]).
A relatively high percentage of respondents stats they did not know what the 
source of their tap water was (13.3%, 1.2% SE). 
The majority of respondents (65.5%) used unfiltered or filtered tap water as 
their main source of drinking water ([@tbl-q5q6]). Bottled or pre-packaged 
water was the primary drinking water sources for 34.2% (1.6% SE) of respondents.
Only 3.1% (0.6% SE) of respondents stated that they had not experienced any
issues with their tap water ([@tbl-q7]). Odd taste (27.5%, 1.4% SE), hardness 
or scale buildup (20.3%, 1.3% SE), and odd smell (18.8%, 1.3% SE) were the most 
common issues. Notably, 14.9% (1.1% SE) of respondent stated they have 
experienced a boil water notice.

```{r}
#| label: tbl-q5q6
#| tbl-cap: The proportion of respondents with the specified household water source and drinking water source.

tar_read(q5_q6_summary)

```




```{r}
#| label: tbl-q7
#| tbl-cap: Proportion of respondents that have experienced specified issues with their drinking water.

tar_read(q7_summary)

```


```{r}
#| label: fig-q11
#| fig-cap: Contaminants of concern


tar_read(q11_summary)$p1
```

We did not observe strong differences among response levels between different
contaminants of potential concern ([@fig-q11]). Pesticides (*extremely 
concerned* = 21%, 1.3% SE; *very concerned* = 21.1%, 1.3% SE), industrial 
chemicals (*extremely concerned* = 20.3%, 1.3% SE; *very concerned* = 20.1%, 
1.3% SE), pathogens (*extremely concerned* = 19.9%, 1.3% SE; *very concerned* = 
20.8%, 1.3% SE), and heavy metals (*extremely concerned* = 19.8%, 1.3% SE; 
*very concerned* = 22.0%, 1.3% SE) are sources that elicited the highest
levels of concern across all respondents. Disinfection byproducts 
(*Not at all concerned* = 22.6%, 1.4% SE; *slightly concerned* = 20.7%, 1.3% SE) 
and PFAS (*not at all concerned* = 23.0%, 1.4% SE; *slightly concerned* = 17.8%, 
1.2% SE) were sources of least concern.



```{r}
#| label: fig-m2
#| fig-cap: Primary drinking water model coefficients (odds-ratios) and 95% CIs. Odd-ratios are relative to unfiltered tap water.
#| fig-asp: 1.1
tar_read(m2_coefs)$p1
```

Based on the multinomial regression model, we found limited effects of 
demographic variables, drinking water sources, and experience with odd taste, 
color, or smell on the odds of using bottled drinking water compared to
unfiltered drinking water. Notably, females had 1.49 times the odds 
(95% CI [1.02, 2.17]) of using bottled drinking water instead of unfiltered 
tap water compared to males. Non-whites had 1.59 times the odds 
(95% CI [1.01, 2.52]) of using bottled drinking water instead of unfiltered 
drinking water compared to white respondents. Renters had 1.64 times the odds
(95% CI [1.07, 2.50]) compared to homeowners.

Conversely, older respondents were more likely to use unfiltered drinking water
compared to filtered water (45:54 age group: OR = 0.29, 95% CI [0.132, 9.636];
55:64 age group: OR = 0.29, 95% CI [0.130, 0.647]; 65+ age group: OR = 0.278, 
95% CI [0.002, 1.25]) but there was not strong evidence for differences
in odds between bottled water and unfiltered drinking water among any of the 
age groups. Education also showed strong effects on the use of filtered drinking
water. Having a bachelor's degree resulted in 4.31 times the odds of using 
filtered drinking water compared to those not graduating high school.
A master's degree had a similarly large effect (OR =  3.06, 95% CI [1.09, 8,59]).

## Sources of information


```{r}
#| label: fig-q12
#| fig-cap: Sources of drinking water quality information.


tar_read(q12_summary)$p1
```

Water providers appear to be the most relied upon source for information on
drinking water with 13.9% (1.1% SE) and 16.4% (1.2% SE) of responses indicating
water providers are *always* or *often* used as a source of information 
([@fig-q12]). Both social media (*always* = 11.5%, ±1% SE; *often* = 15.2%, ±1.1% SE) 
and word-of-mouth (*always* = 9.4%, ±1% SE; *often* = 17.6%, ±1.2% SE) are also frequent
sources of information with similar or higher usage than local news agencies 
(*always* = 8.7%, ±0.9% SE; *often* = 14.3%, ±1.1% SE) or local newspapers
(*always* = 6.6%, ±0.8% SE; *often* = 15.3%, ±1.1% SE). Some of the least
utilized sources included radio or podcasts (*never* = 40.5%, ±1.6% SE; 
*occasionally* = 20.1%, ±1.3% SE), email (*never* = 46.7%, 
±1.6% SE; *occasionally* = 19.6%, ±1.3% SE), national newspapers (*never* = 
51.8%, ±1.6% SE; *occasionally* = 16.7%, ±1.2% SE), and government agencies 
(*never* = 37.5%, ±1.6% SE; *occasionally* = 22.4%, ±1.4% SE).




## Perceived responsibility, trust, and safety



```{r}
#| label: fig-q14
#| fig-cap: Weighted proportional responses for the level of responsibility people believe different entities have for ensuring drinking water safety. Points represent the population-level estimated proportional response, horizontal lines and density ridges respresnt the indicated confidence intervals.
#| fig-asp: 0.8

tar_read(q14_shaded)

```

In general, higher proportions of survey respondents assigned water utilities 
(53.3%, 95% CI [50.1%, 56.5%]), local government (40.7%, 95% CI [37.5%, 43.8%]), 
and state government (36.6%, 95% CI [33.5%, 39.6%]) with the highest levels of 
responsibility for ensuring safety of drinking water ([@fig-q14]).
Moderate responsibility was the most common response for the level of 
responsibility attributed to household residents (26.6%, 95% CI [23.9%, 29.4%]) 
and landlords/property managers (27.6%, 95% CI [24.8%, 30.4%]).
A relatively high proportion of responses indicated residents have no 
responsibility for drinking water quality (16.4%, 95% CI [14.0%, 18.9%]).
Relative to other governmental institutions, fewer people felt the federal
government had full responsibility (27.9%, 95% CI [25.0%, 30.7%]) for drinking 
water safety.



```{r}
#| label: fig-q15
#| fig-cap: Weighted proportional responses for the level of trust people have in the entities responsible for ensuring drinking water safety. Bar lengths represent the mean proportional response with eror bars representing the 95% confidence interval.

tar_read(q15_summary)$p1

```


When asked for their level of trust in these entities for ensuring safety of 
drinking water, 30% (95% CI [27.3%, 33.2%]) of respondents expressed they do not
trust the federal government at all ([@fig-q15]). 
Only 6.5% (95% CI [4.9%, 8.1%]) indicated they fully trust the federal 
government for ensuring drinking water safety.
State government fared slightly better, with 28.7% (95% CI [25.9%, 31.6%]) and 
28.6% (95% CI [25.7%, 31.5%]) of respondents indicating they moderately trust 
or somewhat trust the state government for ensuring safety.
Local and state government had rates of full trust (7.1%, 
95% CI  [5.5%, 8.7%] and 7.6%, 95% CI [5.9%, 9.3%], respectively) comparable to 
the low rate observed for the federal government (6.5%, 95% CI [5.0%, 8.1%]). 

Water utilities appear to be more trusted by the public than other governmental
institutions. While the proportion of responses indicating moderate trust
(27.0%, 95% CI [24.2%, 29.8%]) and mostly trust (22.3%, 95% CI [19.7%, 24.9%])
are comparable to local, state, and federal government, much fewer people
indicated that the "do not trust at all" the water utility (14.9%, 95% CI 
[12.6%, 17.2%]).



```{r}
#| label: fig-m1-m3
#| fig-cap: Estimated effect sizes (odds-ratios) and 95% confidence intervals of selected variables on (1) level of trust in drinking water utilities, and (2) perceived safety of drinking water.
#| fig-asp: 1.1
#| 
tar_read(m1_m3_coefs)
```





The proportional odds model provides evidence that sex/gender, home ownership,
household tap supply, and experience with taste/odor/color issues are associated
with the perception of drinking water safety ([@fig-m1-m3]). 
Females were `r round(1/0.76, 2)` times more likely to rate their drinking
water less safe than males (OR = 0.76, 95% CI [0.59, 0.97]).
Respondents identifying their gender as other had `r round(1/0.13, 2)` times 
the odds (OR = 0.13, 95% CI [0.05, 0.32]) of rating their drinking water less safe
than male respondents.
Home ownership provided a strong signal with renters `r round(1/0.67,2)` times 
more likely to provide a lower drinking water safety ranking than homeowners 
(OR = 0.67, 95% CI [0.51, 0.89]).
Respondents on private water supplies had 1.54 times that odds of providing
a higher drinking water safety rating than people on municipal supplies 
(OR = 1.54, 95% CI [1.10, 2.16]). 
People that experienced taste/odor/color issues with `r round(1/0.32, 2)` times
more likely to provide a lower safety rating than those that never experienced 
an issue (OR = 0.32, 95% CI [0.24, 0.41]).

We identified difference among some age groups, but a clear pattern was not
evident. Respondents in the 45-54 age bracket were `r round(1/0.45, 2)`
times more likely to provide a lower drinking water safety rating than the
18-24 year old age group.



# Discussion

# Notes

@switzerClassRaceEthnicity2018 finds interaction between race and other socioeconomic
factors particularly important. For communities of lower socioeconomic status, race plays 
an even larger role in predicting the likelihood of SDW violations. This may suggest
non-white poor communities have higher propensity for distrust in water quality,
but we fail to incorporate these interactions in our model due to sample sizes. 


# Acknowledgements

TBD

# Author Contributions

AB: Conceptualization, Investigation, Project administration, Supervision, Writing - review & editing.
AM: Data curation, investigation, methodology, writing - original draft.
SD: Investigation, methodology, writing - original draft.
MS: Methodology, formal analysis, software, visualization, writing - original draft.

